Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-09 paper.qmd
Removed long comments

<!-- - What's the problem:

  Non-linear dimension reduction being used to summarise high-dimensional data.

  - Summary of literature

  Relevant high-d vis, NLDR history
-->

@@

<!-- Linked brushing as done by @article21

- Model-in-the-data-space: how can we represent the model, eg plane for PCA, grid of values for classification boundaries, ellipses for LDA and mclust, nets for SOM.-->

@@

At first glance, thinking of NLDR as a modeling technique might seem strange. It is a simplified representation or abstraction of a system, process, or phenomenon in the real world. The \pD{}   observations are the realization of the phenomenon, and the \kD{}   NLDR layout is the simplified representation. From a statistical perspective we can consider the distances between points in the \kD{}   layout to be variance that the model explains, and the (relative) difference with their distances in \pD{}   is the error, or unexplained variance. We can also imagine that the positioning of points in \gD{}    represent the fitted values, that will have some prescribed position in \pD{}   that can be compared with their observed values. This is the conceptual framework underlying the more formal versions of factor analysis [@cfa69] and MDS. (Note that, for this thinking the full \pD{}   data needs to be available, not just the interpoint distances.)
<!--### Notation -->

<!-- @tbl-notation summarises the notation used to explain the new methodology. The observed data is denoted as $\mathbfit{x}_{n \times p}$ where $x_{ij}$ would indicate the $i^{th}$ observation on the $j^{th}$ variable sampled from a population $\mathbfit{X}$. To refer to variable $j$, we would use $X_j$.-->

<!--
$X_{n \times p} = \begin{bmatrix} \textbf{x} _{1} & \textbf{x}_ {2} & \cdots & \textbf{x}_{n} \\  \end{bmatrix}^\top$

$Y_{n \times d} = \begin{bmatrix} \textbf{y} _{1} & \textbf{y}_ {2} & \cdots & \textbf{y}_{n} \\  \end{bmatrix}^\top$

$C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2})$

$C_k^{(p)} \equiv (C_{kx_1}, ..., C_{kx_p})$ $p$-D mappings of 2D hexagon bin centroids of the $k^{th}$ hexagon

-->

@@

Because we are working with distances between points, starting with data having a standard scale, e.g. [0, 1], is recommended. The default should take the aspect ratio produced by the NLDR $(r_1, r_2, ..., r_k)$ into account. When $k=2$, as in hexagon binning, the default range is $[0, y_{i,\text{max}}], i=1,2$, where $y_{1,\text{max}}=1$ and $y_{2,\text{max}} = r_2/r_1$ (@fig-NLDR-two-curvy). If the NLDR aspect ratio is ignored then set $y_ {2,\text{max}} = 1.$ <!-- \times \frac{2}{\sqrt{3}}$. (The $\frac{2}{\sqrt{3}}$ accounts for the different height ($a_1$) and width ($a_2$) of a regular hexagon.) The scaling of data should take the size of the hexagons into account, but choice of number of bins should. -->

@@

<!-- Number of bins is set by fixing $b_1$, which determines the binwidth accounting the offset $q_1$ and breaks on $x_1$, is calculated as  -->

<!-- $$ -->
<!-- a_1 = \frac{r_1 + q_1}{b_1}. -->
<!-- $$ -->

<!-- The computed binwidth then determines the number of vertical bins, $b_2$, and vertical binwidth, which are computed based on $r_2$, the offset $q_2$, and the height/width ratio of a regular hexagon, $\frac{2}{\sqrt{3}}$ as -->

<!-- $$ -->
<!-- a_2 = \frac{r_2 + q_2}{0.75 \times b_2}. -->
<!-- $$ -->

<!-- To define a hexagon grid across the \kD{} space, with hexagons of fixed height ($a_1$), width ($a_2$),  it is necessary to determine how many hexagons should be represented along each axis in the \kD{} space.

When $k=2$, the number of bins along the $x$ and $y$ axes, $b_1$ and $b_2$, is computed by considering the scaling factors ${r_1, r_2}$, along with the offset along the axes ($q_1$, $q_2$), and the height ($a_1$) and width ($a_2$) of a regular hexagon (@fig-NLDR-two-curvy (b)). ($b_1 = \frac{r_1 + q_1}{a_2}$, and $b_2 = \frac{r_2 + q_2}{0.75 \times a_1}$) where $0.75 \times a_1$ accommodate to have hexagons fill the entire 2D space without leaving any gaps between them.)-->

<!-- XXX We don't need parameters for the regular hexagon, these are fixed constants. -->

@@

<!-- Points are allocated to the bin they fall into based on the nearest centroid. In situations where a point is equidistant from multiple centroids, tie-breaking rules are applied. If multiple centroids are in the same row, the point is assigned to the leftmost centroid. If multiple centroids are in different rows, the point is assigned to the bottom centroid. -->

<!-- $\{ i \in H_h, h = 1, \dots, b, \text{ and } i = 1, \dots, n\}$ -->

@@

<!--
Define a hexagon grid across the \kD{} space, using hexagons with fixed height ($a_2$), width ($a_1$). Each of the \kD{} points will belong to a hexagon bin. That is, for each $y \in \mathbfit{Y}$, we can (uniquely) identify the hexagon that the point belongs to. This identification is done finding the nearest bin centroid for the \kD{} points by considering the \kD{} Euclidean distance.

When $k=2$, the starting coordinates $(s_1, s_2)$ mark the lower left of the grid. This is the bottom left bin centroid. By starting from there, points are generated to fill the grid accounting $b_1$, $b_2$, $a_1$, and $a_2$. 
-->

<!--We deliberately separate out the creation of the hexagon grid from the mapping of points on the grid.-->

@@

<!-- When $k = 2$ Delaunay triangulation on $C^{(2)}$ generates the model in \gD{} space, which is a triangular mesh (@fig-NLDR-two-curvy (d)). It generates convex hulls of $C^{(2)}$ such that the circumcircle of every triangle in the triangulation contains no other points from $C^{(2)}$. -->

@@

<!--need to add what is meant by a long edge-->

@@

<!-- XXX You need to have consistent colouring. In langevitour the points corresponding to observations are purple. They should be purple in the \gD{} view too.  -->

<!-- XXX Should there be a second long edge removal? You can compute the edges in \pD{} now, and if there are long edges maybe these should be pruned, and we re-plot the \gD{} view.  -->

<!-- langevitour with pD model for two-curvy clusters -->
```{r}
#| label: best-fit-tsne
#| fig-cap: "Lifting the fitted model into \\pD{}. Starting from a default tSNE layout (a) of non-linear clusters data ($n =  2000$ and $p = 4$), from which the \\gD{} wireframe is constructed (a) and then lifted to $4\\text{-}D$ (b, c). The fit is reasonbly tight with the data, although it does not fully spread the full width. This highlights a key characteristic of tSNE, which compresses the data when applying tSNE. Furthermore, the sparse space observed in the middle of the tSNE layout is a result of (hyper-)parameter choice. Video of the langevitour animation is available at <>."
#| fig-width: 8
#| fig-height: 2

@@

<!--The prediction approach involves finding the nearest \kD{} model point for a new \pD{} point. We define the function $z: \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{m\times p}$, where $z(i) = \arg\min_{j = 1, \dots, b} \sqrt{\sum_{v=1}^{p}(x_{iv} - C^{(p)}_{jv})^2}$ maps each \pD{} point to its nearest \pD{} mapping of the model. Therefore, the new observation $i$ falls in the hexagon, $H_h = \{i| z(i) = h\}$ and the corresponding \kD{} bin centroids, $C_{h}^{(2)} = (c_{h1}, c_{h2})$ be the predicted values.-->

<!-- ::: {#fig-two_curvy-pred-sc layout-ncol="2" fig-pos="H"} -->
<!-- ![](figures/two_curvy/sc_true_only.png){width="150" fig-align="center"} -->

<!-- ![](figures/two_curvy/pred_true_view.png){width="150" fig-align="center"} -->

<!-- A view of the true model in projections from $7\text{-}D$, and predictions of the true model in \gD{}, for the S-curve data. The predictions fits the UMAP layout which means that it capture the geometry of S-curve with UMAP. -->
<!-- ::: -->

<!-- XXX Prediction should have a jitter option, to spread the predicted points out fully in the hexagons. -->

### Tuning
<!-- removal of low density bins, removing long edges, choice of bins-->

@@

<!--The starting position of the hexagonal grid is important because different starting points result in different distributions of data across bins, even with the same total number of bins. This variation affects the model due to the differing number of non-empty bins. Therefore, it is necessary to evaluate various starting points with different total numbers of bins to determine which configurations are more effective at capturing the structure and fitting the model.-->

@@

<!-- To determine the effective $b$, candidate values are selected based on the range between the minimum and approximate maximum $b_1$, because $b_2$ is computed from $b_1$. The minimum $b_1$ is set to $2$, while the maximum number is estimated by taking the square root of $\frac{n}{2}$. By evaluating MSE across varying $b$ within this range for different $q$, helps to determine an appropriate values for $b$ and $q$ (@fig-param-two_curvy (a)).-->

<!--add MSE vs total number of error plot-->

<!--To generate errors for different total number of bins-->

#### Measurement of capturing the data shape in 2-D

The area of a hexagon is defined as $A = 3\sqrt{3}l^2/2$ where $l$ is the side length of the hexagon. If we know $a_1$ and $a_2$, $l$ can be computed (see appendix). The density of a hexagon grid is calculated as $\sum^{h}_{i=1}n_h/A$ and the proportion is $\sum^{h}_{i=1}n_h/Ab$. The baseline proportion is the proportion density at the smallest possible value of $a_1$. The relative proportion density is the ratio of the observed proportion density to the baseline proportion density.

@@

<!-- Once setting up the hexagon grid with an appropriate number of bins, some hexagon bins may have few or no data points within them (@fig-bins-two_curvy (b)). To ensure comprehensive coverage of the NLDR data, it is necessary to select hexagon bins with a considerable number of data points. This involves calculating the number of points within each hexagon. Then, the standard count is computed by dividing the number of points within each hexagon by the maximum number of points in the grid. Next, bins with a standard count less than a benchmark value are removed (@fig-param-two_curvy (c)). There is no specific rule for selecting a benchmark value. However, the following steps can help determine a suitable value for removing low-density hexagons:

1. Plot the distribution of the standardized counts (@fig-param-two_curvy (b)).
2. Examine the distribution of counts.
3. Select the first quantile value if the distribution is skewed.
-->

@@

<!--add MSE vs density (0 to 1)-->

<!--base line -->

<!-- Furthermore, selecting the benchmark value for removing low-density hexagons is important. Removing unnecessary bins may lead to the formation of long edges and an uneven \gD{} model. Hence, rather than solely relying on the benchmark value to identify hexagons for removal, it's essential to consider the standard number of points in the neighboring hexagons of the identified low-density bins (see @fig-lwd-two_curvy (b)). If neighboring bins also show low counts, only those bins will be removed. The remaining bins are used to construct the \gD{} model.    -->

@@

<!-- #### Removing long edges -->

<!-- Edges define the neighbourhood structure, in order to provide a smooth \gD{} representation of the fitted model. @fig-two-curvy-true-proj shows a wire frame of the true model that was used to generate the non-linear clusters example data. The ideal is that the representation of the fitted model, at least for this example where we know the true model, should look similar to this.  -->

<!-- The Delaunay triangulation will ensure that all centroids are connected into a triangular mesh. For some structures, like clustered data, or highly non-linear shapes, breaks in the mesh are meaningful. When separated clusters are present the mesh should be broken across the gaps. For non-linear structures like the C-shaped curvilinear, the mesh should run unbroken along the C, but there should be no edges connecting the top of the C directly to the bottom of the C. For these reasons it is necessary to remove edges from the mesh in some applications. -->

<!-- The decision on edge length removal is made based on the distribution of edge lengths. In particular, a gap between values, where there a concentration of small values and then a few larger values, likely suggests a cutoff for edge removal. Because the triangulation is typically done on the hexagon centroids, there are particular discrete edge lengths, based on bin widths. @fig-rm-lg illustrates edge length distributions.  -->

<!-- There is an additional step that is needed. When the model is lifted into \pD{}, if the fit is good all the edges should be relatively small in this space, too. If this is not the case, then there are several possible actions: (1) re-do the NLDR to get a more representative layout; (2) identify the edge and remove it from the model, in \gD{} and \pD{}; (3) consider different values for the model fit, number of bins, initial bin position or removing low density bins. -->

<!-- ```{r} -->
<!-- #| label: model-two-curvy-umap -->

<!-- umap_two_curvy <- read_rds(file = "data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_umap_n-neigbors_15_min-dist_0.1.rds") -->

<!-- num_bins_x_two_curvy <- 42 -->

<!-- algo_obj_two_curvy3 <- gen_nldr_vis_algo_obj( -->
<!--   high_d_data = training_data_two_curvy,  -->
<!--   nldr_data = umap_two_curvy,  -->
<!--   num_x_bins = num_bins_x_two_curvy) -->

<!-- umap_two_curvy_scaled <- algo_obj_two_curvy3$nldr_scaled -->
<!-- distance_two_curvy <- algo_obj_two_curvy3$distance_df -->
<!-- tr_from_to_df_two_curvy <- algo_obj_two_curvy3$tr_from_to_df -->
<!-- benchmark_two_curvy <- algo_obj_two_curvy3$benchmark -->
<!-- df_bin_centroids_two_curvy <- algo_obj_two_curvy3$df_bin_centroids -->
<!-- df_bin_two_curvy <- algo_obj_two_curvy3$df_bin -->
<!-- hex_grid_with_counts <- algo_obj_two_curvy3$hex_grid_with_counts -->
<!-- a1 <- algo_obj_two_curvy3$a1 -->

<!-- trimesh_two_curvy_umap <- ggplot() + -->
<!--   geom_segment(data = tr_from_to_df_two_curvy, -->
<!--                aes( -->
<!--                  x = x_from, -->
<!--                  y = y_from, -->
<!--                  xend = x_to, -->
<!--                  yend = y_to), -->
<!--                colour = "#000000", -->
<!--                alpha = 0.5, -->
<!--                linewidth = 0.5) + -->
<!--   geom_point(data = umap_two_curvy_scaled, -->
<!--              aes( -->
<!--                x = UMAP1, -->
<!--                y = UMAP2 -->
<!--              ), -->
<!--              color = "#636363", -->
<!--              alpha = 0.5, -->
<!--              size = 0.5 -->
<!--   ) + -->
<!--   coord_equal() + -->
<!--   xlim(c(0, 1)) + -->
<!--   ylim(c(0, 1)) + -->
<!--   interior_annotation("a1", sc_ltr_pos, cex = 2) -->

<!-- ``` -->

<!-- ```{r} -->

<!-- # With the computed benchmark value -->
<!-- # distance_df_small_edges_two_curvy <- distance_two_curvy |> -->
<!-- #   filter(distance < benchmark_two_curvy) -->

<!-- ####With diff benchmark values -->

<!-- ## Benchmark 2 -->

<!-- benchmark_two_curvy <- 8 * a1 -->

<!-- distance_df_small_edges_two_curvy3 <- distance_two_curvy |> -->
<!--   filter(distance < benchmark_two_curvy) -->

<!-- tr_from_to_df_two_curvy4 <- inner_join( -->
<!--   tr_from_to_df_two_curvy, distance_df_small_edges_two_curvy3, -->
<!--   by = c("from", "to")) -->

<!-- trimesh_two_curvy_removed2_umap <- ggplot() + -->
<!--   geom_segment(data = tr_from_to_df_two_curvy4, -->
<!--                aes( -->
<!--                  x = x_from, -->
<!--                  y = y_from, -->
<!--                  xend = x_to, -->
<!--                  yend = y_to), -->
<!--                colour = "#000000", -->
<!--                alpha = 0.5, -->
<!--                linewidth = 0.5)+ -->
<!--   geom_point(data = umap_two_curvy_scaled, -->
<!--              aes( -->
<!--                x = UMAP1, -->
<!--                y = UMAP2 -->
<!--              ), -->
<!--              color = "#636363", -->
<!--              alpha = 0.5, -->
<!--              size = 0.5 -->
<!--   ) + -->
<!--   coord_equal() + -->
<!--   xlim(c(0, 1)) + -->
<!--   ylim(c(0, 1)) + -->
<!--   interior_annotation("b1", sc_ltr_pos, cex = 2) -->

<!-- ## Benchmark 1 -->

<!-- benchmark_two_curvy <- 2.5 * bin_width -->

<!-- distance_df_small_edges_two_curvy2 <- distance_two_curvy |> -->
<!--   filter(distance < benchmark_two_curvy) ## 0.231 -->

<!-- tr_from_to_df_two_curvy3 <- inner_join( -->
<!--   tr_from_to_df_two_curvy, distance_df_small_edges_two_curvy2, -->
<!--   by = c("from", "to")) -->

<!-- trimesh_two_curvy_removed1_umap <- ggplot() + -->
<!--   geom_segment(data = tr_from_to_df_two_curvy3, -->
<!--                aes( -->
<!--                  x = x_from, -->
<!--                  y = y_from, -->
<!--                  xend = x_to, -->
<!--                  yend = y_to), -->
<!--                colour = "#000000", -->
<!--                alpha = 0.5, -->
<!--                linewidth = 0.5)+ -->
<!--   geom_point(data = umap_two_curvy_scaled, -->
<!--              aes( -->
<!--                x = UMAP1, -->
<!--                y = UMAP2 -->
<!--              ), -->
<!--              color = "#636363", -->
<!--              alpha = 0.5, -->
<!--              size = 0.5 -->
<!--   ) + -->
<!--   coord_equal() + -->
<!--   xlim(c(0, 1)) + -->
<!--   ylim(c(0, 1)) + -->
<!--   interior_annotation("c1", sc_ltr_pos, cex = 2) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #| label: compare-2d-and-4d-dist -->

<!-- ### Compute highD distance -->
<!-- dist_vec <- proxy::dist(x = df_bin_two_curvy[, -1], method = "Euclidean") |> as.vector() -->

<!-- from_vec <- c() -->
<!-- to_vec <- c() -->
<!-- num_obs <- 1:(NROW(df_bin_two_curvy) - 1) -->

<!-- for (obs in num_obs) { -->

<!--   from_val <- rep(obs, (NROW(df_bin_two_curvy) - obs)) -->
<!--   if ((obs + 1) <= NROW(df_bin_two_curvy)) { -->
<!--     to_val <- (obs + 1):NROW(df_bin_two_curvy) -->
<!--   } -->
<!--   from_vec <- append(from_vec, from_val) -->
<!--   to_vec <- append(to_vec, to_val) -->

<!-- } -->

<!-- dist_highd <- tibble::tibble(from = from_vec, to = to_vec, dist_highd = dist_vec) -->

<!-- ## To plot the distribution of distances -->

<!-- # Define bin width and starting point -->
<!-- start_point <- a1/2 -->

<!-- # Create bins and calculate the mean value for each bin range -->
<!-- bins <- seq(start_point, max(distance_two_curvy$distance) + a1, -->
<!--             by = a1) -->
<!-- bin_labels <- bins[-length(bins)] + a1 / 2  # mean of each bin range -->

<!-- distance_two_curvy_dist <- distance_two_curvy |> -->
<!--   mutate(bin = cut(distance, breaks = bins, include.lowest = TRUE, labels = bin_labels)) -->

<!-- distance_two_curvy_dist <- left_join(distance_two_curvy_dist, dist_highd, by = c("from", "to")) -->

<!-- text_df <- tibble::tibble( -->
<!--   x = c(0.12, 0.43), -->
<!--   y = c(10, 10), -->
<!--   text = c("c", "b") -->
<!-- ) -->

<!-- distance_points_umap <- ggplot( -->
<!--   distance_two_curvy_dist, -->
<!--   aes(x = distance, -->
<!--       y = dist_highd)) + -->
<!--   geom_point(alpha = 0.5) + -->
<!--   geom_text( -->
<!--     data=text_df, -->
<!--     aes(x=x, y=y, -->
<!--         label=text), -->
<!--     colour="#bdbdbd", -->
<!--     size = 10) + -->
<!--   geom_vline(xintercept = 8 * bin_width, -->
<!--              linetype="solid", -->
<!--              color = "#bdbdbd", -->
<!--              linewidth=1) + -->
<!--   geom_vline(xintercept = 2.5 * bin_width, -->
<!--              linetype="solid", -->
<!--              color = "#bdbdbd", -->
<!--              linewidth=1) + -->
<!--   ylab(expression(d^{(4)})) + -->
<!--   xlab(expression(d^{(2)})) + -->
<!--   theme_minimal() + -->
<!--   theme(aspect.ratio = 0.75, -->
<!--         panel.border = element_rect(fill = 'transparent'), -->
<!--         plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5), -->
<!--         panel.grid.major.x = element_blank(), -->
<!--         axis.ticks.x = element_line(), -->
<!--         axis.ticks.y = element_line(), -->
<!--         axis.title = element_text(size = 15), -->
<!--         axis.text = element_text(size = 12)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #| label: proj-tsne-model -->

<!-- df_bin_two_curvy <- df_bin_two_curvy |> -->
<!--   select(-hexID) |> -->
<!--   mutate(type = "model") -->

<!-- # Apply the scaling -->
<!-- df_model_data_two_curvy <- bind_rows(data_two_curvy, df_bin_two_curvy) -->

<!-- scaled_two_curvy <- scale_data_manual(df_model_data_two_curvy, "type") |> -->
<!--   as_tibble() -->

<!-- scaled_two_curvy_data <- scaled_two_curvy |> -->
<!--   filter(type == "data") |> -->
<!--   select(-type) -->

<!-- scaled_two_curvy_data_model <- scaled_two_curvy |> -->
<!--   filter(type == "model") |> -->
<!--   select(-type) -->

<!-- ## First projection -->
<!-- projection <- cbind( -->
<!--   c(0.37847,-0.06517,0.04231,0.01655), -->
<!--   c(-0.06466,-0.27052,0.22131,-0.15235)) -->

<!-- proj_obj1 <- get_projection(projection = projection,  -->
<!--                             proj_scale = 1,  -->
<!--                             scaled_data = scaled_two_curvy_data,  -->
<!--                             scaled_data_model = scaled_two_curvy_data_model,  -->
<!--                             distance_df_small_edges = distance_df_small_edges_two_curvy2,  -->
<!--                             axis_param = list(limits = 0.6,  -->
<!--                                               axis_scaled = 2, -->
<!--                                               axis_pos_x = -0.5,  -->
<!--                                               axis_pos_y = -0.5,  -->
<!--                                               threshold = 0.04)) -->

<!-- ## Projection with Delauany triangulation -->

<!-- projected_model <- as.matrix(scaled_two_curvy_data_model) %*% projection_scaled -->

<!-- projected_model_df <- projected_model |> -->
<!--   tibble::as_tibble(.name_repair = "unique") |> -->
<!--   dplyr::rename(c("proj1" = "...1", -->
<!--                   "proj2" = "...2")) |> -->
<!--   dplyr::mutate(ID = dplyr::row_number()) -->

<!-- model_df1 <- dplyr::left_join( -->
<!--   distance_two_curvy |> select(-distance), -->
<!--   projected_model_df, -->
<!--   by = c("from" = "ID")) -->

<!-- names(model_df1)[3:NCOL(model_df1)] <- paste0(names(projected_model_df)[-NCOL(projected_model_df)], "_from") -->

<!-- model_df1 <- dplyr::left_join(model_df1, projected_model_df, by = c("to" = "ID")) -->
<!-- names(model_df1)[(2 + NCOL(projected_model_df)):NCOL(model_df1)] <- paste0(names(projected_model_df)[-NCOL(projected_model_df)], "_to") -->

<!-- ## With benchmark option 2 -->

<!-- model_df2 <- dplyr::left_join( -->
<!--   distance_df_small_edges_two_curvy3 |> select(-distance), -->
<!--   projected_model_df, -->
<!--   by = c("from" = "ID")) -->

<!-- names(model_df2)[3:NCOL(model_df2)] <- paste0(names(projected_model_df)[-NCOL(projected_model_df)], "_from") -->

<!-- model_df2 <- dplyr::left_join(model_df2, projected_model_df, by = c("to" = "ID")) -->
<!-- names(model_df2)[(2 + NCOL(projected_model_df)):NCOL(model_df2)] <- paste0(names(projected_model_df)[-NCOL(projected_model_df)], "_to") -->

<!-- two_curvy_proj_first_model1_umap <- plot_proj( -->
<!--     proj_obj = proj_obj1,  -->
<!--     point_param = c(0.5, 0.2, "#636363"), # size, alpha, color -->
<!--     line_param = c(0.5, 0.5, "#000000"), # linewidth, alpha -->
<!--     plot_limits = c(-0.6, 0.6),  -->
<!--     title = "c2",  -->
<!--     cex = 2,  -->
<!--     position = sc_ltr_pos, -->
<!--     axis_text_size = 4,  -->
<!--     is_category = FALSE) + -->
<!--     theme( -->
<!--         legend.position = "none" -->
<!--     ) -->

<!-- ## Projection with Delauany triangulation -->

<!-- proj_obj1_temp1 <- proj_obj1 -->
<!-- proj_obj1_temp2 <- proj_obj1 -->
<!-- proj_obj1_temp1[["model_df"]] <- model_df1 -->

<!-- two_curvy_proj_model_delaunay_umap <- plot_proj( -->
<!--     proj_obj = proj_obj1_temp1,  -->
<!--     point_param = c(0.5, 0.2, "#636363"), # size, alpha, color -->
<!--     line_param = c(0.5, 0.5, "#000000"), # linewidth, alpha -->
<!--     plot_limits = c(-0.6, 0.6),  -->
<!--     title = "a2",  -->
<!--     cex = 2,  -->
<!--     position = sc_ltr_pos, -->
<!--     axis_text_size = 4,  -->
<!--     is_category = FALSE) + -->
<!--     theme(aspect.ratio = 1,  -->
<!--           legend.position = "none") -->

<!-- proj_obj1_temp2[["model_df"]] <- model_df2 -->

<!-- two_curvy_proj_model_benchmark2_umap <- plot_proj( -->
<!--     proj_obj = proj_obj1_temp2,  -->
<!--     point_param = c(0.5, 0.2, "#636363"), # size, alpha, color -->
<!--     line_param = c(0.5, 0.5, "#000000"), # linewidth, alpha -->
<!--     plot_limits = c(-0.6, 0.6),  -->
<!--     title = "b2",  -->
<!--     cex = 2,  -->
<!--     position = sc_ltr_pos, -->
<!--     axis_text_size = 4,  -->
<!--     is_category = FALSE) + -->
<!--     theme(aspect.ratio = 1,  -->
<!--           legend.position = "none") -->

<!-- ``` -->

<!-- ```{r} -->
<!-- #| label: fig-rm-lg -->
<!-- #| fig-cap: "The choice of long edge removal significantly impacts the model's accuracy. \\gD{} Vs $7\\text{-}D$ distances between the bin centroids, where the model is fitted with UMAP for the non-linear clusters data. The two benchmarks for long edge removal: (b) benchmark = 8$a_1$ and (c) benchmark = 2.5$a_1$, where $a_1$ = 0.03 are made. The wireframes in \\gD{} and a projection from $7\\text{-}D$ are shown with Delaunay triangulation (a1, a2) and the two benchmark values (b1, b2, c1, c2). With the benchmark set at 6.5$a_1$, long edges exist within the C-shaped cluster, which distrupts the model. In contrast, using 2$a_1$ is a better option, as it effectively captures the non-linear shapes of the clusters. To generate an accurate model that captures the data structure, the choice of long edge removal does matter." -->
<!-- #| fig-height: 15 -->
<!-- #| fig-width: 15 -->

<!-- free(distance_points_umap) +  -->
<!--     wrap_plots(trimesh_two_curvy_umap, trimesh_two_curvy_removed2_umap, -->
<!--                trimesh_two_curvy_removed1_umap, two_curvy_proj_model_delaunay_umap, -->
<!--                two_curvy_proj_model_benchmark2_umap, two_curvy_proj_first_model1_umap, -->
<!--                ncol = 3) + -->
<!--     plot_layout(heights = c(0.8, 1)) -->
<!-- ``` -->

@@

<!-- Brushing the high model error in the \gD{} layout, with dense C-shaped data explained in @sec-effect-dens, helps in locating the positions of the related \pD{} points in langevitour view. This process helps in identifying the high model errors that occur due to the sparse end (add the video). -->

<!-- Furthermore, when examining the digit 1 from the MNIST dataset (@sec-mnist), brushing the small cluster in the \gD{} layout helps in locating the positions of the related \pD{} points in langevitour view. This process helps in identifying how close the small cluster to the big cluster (add the video). We can also explore which data corresponds to the twisted pattern of the big cluster (add the video).  -->

@@

The models are fitted to each these layouts. @fig-five-gau-projs a2, b2, c2 show the fitted models in a projection of the \fD{} space, taken from a tour. These clusters are fully \fD{} in nature, so we would expect the model to be a *crumpled sheet* that stretches in all four dimensions. This is what is mostly observed for tSNE and UMAP. The curious detail is that the model for PaCMAP is closer to a *pancake* in shape in every cluster! This seingle projection only shows this in three of the five clusters but if we examine a different projection the other clusters exhibit the pancake also. While we don't know what exactly causes this it is likely due to some ordering of points in the \gD{} PaCMAP layout that induces the flat model. One could imagine that if the method used principal components on all the data, that it might induce some ordering that would produce the flat model. If this were the reason, the pancaking would be the same in all clusters, but it is not: The pancake is visible in some clusters in some projections but in other clusters it is visible in different projections. It might be due to some ordering by nearest neighbors in a cluster. The PaCMAP documentation doesn't provide any helpful clues. That this happens, though, makes the PaCMAP layout inadequate for representing the high-dimensional data. <!-- likely that the points are organised by the global principal components. So the layout of the model in \fD{} doesn't reflect the dimension of each cluster. This can be happen because PaCMAP balances three components (near, mid-range, and far) rather than focusing solely on near neighbors. Mid-range neighbors ensure that transitions between clusters or groups are preserved and less concern with local preservation.-->

@@

<!-- XXX WHY THESE EXAMPLES. NEEDS AN OVERVIEW PARAGRAPH -->

@@

### Single-cell gene expression {#sec-pbmc}
<!--
- NLDR view used to illustrate clusters
- Use our method to assess is it a reasonable representation
- Demonstrate that it is not
- Illustrate how to use our method to get a better representation
-->

<!-- XXX What is single cell data???? -->

@@

<!-- As shown in @fig-NLDR-variety, layout c, d, and e that show small separations between clusters are universally optimal. But layout d with little separations perform well at tiny binwidth and poorly as binwidth increases. Also, layout e performs well at large binwidths. The layout c is chosen as the best, since it used the UMAP with a different (hyper-) parameter settings. -->

<!-- The visualization of the model generated for the selected layout in $9\text{-}D$ shows edges between the clusters (@fig-mnist-tri-proj (b2)). This provides evidence for the small separations between clusters. Furthermore, the data does not uniformally distributed along the clusters and therefore shows densed points (@fig-mnist-tri-proj (b2). Also, the clusters are non-linear shaped (@fig-mnist-tri-proj (b3).   -->

<!-- To determine whether the UMAP representation with the specific (hyper-)parameter choice suggested by @chen2023 preserves the data structure present in $9\text{-}D$ (@fig-model-pbmc-author-proj (b)), we visualize the model constructed with UMAP overlaid on the $9\text{-}D$ data.  -->

<!-- But, when visualizing the *model-in-the-data-space* some unobserved structures can be seen. Some clusters have non-linear continuity patterns and high-density patches (@fig-model-pbmc-author-proj). -->

<!-- As shown in @fig-NLDR-variety (e), there are three well-separated clusters, although the separation between the clusters are small. Additionally, non-linear structures can also be observed within the clusters (@fig-NLDR-variety (e)). This demonstrates that tSNE accurately captures the data structure of the PBMC3k dataset, which UMAP did not achieve. -->

<!-- In order to find a reasonable NLDR representation for the PBMC3k dataset, the MSE for different $b_1$ using UMAP with different (hyper-)parameter settings and tSNE with default (hyper-)parameter setting (@fig-pbmc-mse) were calculated. After analyzing the results, it was found that tSNE with default (hyper-)parameter setting (perplexity: $30$) achieved the lowest error. Therefore, tSNE with a perplexity value set to $30$, the default parameter setting, is considered as a reasonable representation for the PBMC3k dataset. -->

>>>>

@@

@@

At first glance, thinking of NLDR as a modeling technique might seem strange. It is a simplified representation or abstraction of a system, process, or phenomenon in the real world. The \pD{}   observations are the realization of the phenomenon, and the \kD{}   NLDR layout is the simplified representation. From a statistical perspective we can consider the distances between points in the \kD{}   layout to be variance that the model explains, and the (relative) difference with their distances in \pD{}   is the error, or unexplained variance. We can also imagine that the positioning of points in \gD{}    represent the fitted values, that will have some prescribed position in \pD{}   that can be compared with their observed values. This is the conceptual framework underlying the more formal versions of factor analysis [@cfa69] and MDS. (Note that, for this thinking the full \pD{}   data needs to be available, not just the interpoint distances.)

@@

Because we are working with distances between points, starting with data having a standard scale, e.g. [0, 1], is recommended. The default should take the aspect ratio produced by the NLDR $(r_1, r_2, ..., r_k)$ into account. When $k=2$, as in hexagon binning, the default range is $[0, y_{i,\text{max}}], i=1,2$, where $y_{1,\text{max}}=1$ and $y_{2,\text{max}} = r_2/r_1$ (@fig-NLDR-two-curvy). If the NLDR aspect ratio is ignored then set $y_ {2,\text{max}} = 1.$

@@

@@

@@

@@

@@

@@

```{r}
#| label: best-fit-tsne
#| fig-cap: "Lifting the fitted model into \\pD{}. Starting from a default tSNE layout (a) of non-linear clusters data ($n =  2000$ and $p = 4$), from which the \\gD{} wireframe is constructed (a) and then lifted to $4\\text{-}D$ (b, c). The fit is reasonbly tight with the data, although it does not fully spread the full width. This highlights a key characteristic of tSNE, which compresses the data when applying tSNE. Furthermore, the sparse space observed in the middle of the tSNE layout is a result of (hyper-)parameter choice. Video of the langevitour animation is available at <>."
#| fig-width: 8
#| fig-height: 2

@@

### Tuning

@@

@@

#### Measurement of capturing the data shape in 2-D 
The area of a hexagon is defined as $A = 3\sqrt{3}l^2/2$ where $l$ is the side length of the hexagon. If we know $a_1$ and $a_2$, $l$ can be computed (see appendix). The density of a hexagon grid is calculated as $\sum^{h}_{i=1}n_h/A$ and the proportion is $\sum^{h}_{i=1}n_h/Ab$. The baseline proportion is the proportion density at the smallest possible value of $a_1$. The relative proportion density is the ratio of the observed proportion density to the baseline proportion density.

@@

@@

@@

@@

@@

The models are fitted to each these layouts. @fig-five-gau-projs a2, b2, c2 show the fitted models in a projection of the \fD{} space, taken from a tour. These clusters are fully \fD{} in nature, so we would expect the model to be a *crumpled sheet* that stretches in all four dimensions. This is what is mostly observed for tSNE and UMAP. The curious detail is that the model for PaCMAP is closer to a *pancake* in shape in every cluster! This seingle projection only shows this in three of the five clusters but if we examine a different projection the other clusters exhibit the pancake also. While we don't know what exactly causes this it is likely due to some ordering of points in the \gD{} PaCMAP layout that induces the flat model. One could imagine that if the method used principal components on all the data, that it might induce some ordering that would produce the flat model. If this were the reason, the pancaking would be the same in all clusters, but it is not: The pancake is visible in some clusters in some projections but in other clusters it is visible in different projections. It might be due to some ordering by nearest neighbors in a cluster. The PaCMAP documentation doesn't provide any helpful clues. That this happens, though, makes the PaCMAP layout inadequate for representing the high-dimensional data.

@@

@@

### Single-cell gene expression {#sec-pbmc}

@@
